Miniproject - Introduction to Schemes

### 2022-03-31 Schemes/submission.tex
Shrunk diagram to make it fit the page
Added introduction and conlusion to schemes

\tableofcontents

@@

    Consider $\p \in \D{f} \cap \D{g}$, so $f, g \notin \p$.
    Since $\p$ is prime, it follows that $fg \notin \p$ and so $\D{fg} \subseteq \D{f} \cap \D{g}$ is an open neighborhood of $\p$.
\end{proof}
\begin{lemma}
    \label{prop:homogeneous_part_prime}
    Let $\p \leq S$ be a prime ideal.
    Then
    \begin{equation*}
        \p' := \langle f \in \p \ | \ \text{$f$ homogeneous} \rangle \leq S
    \end{equation*}
    is a (homogeneous) prime ideal.
\end{lemma}
\begin{proof}
    Consider $f, g \in S$ with $f g \in \p'$ and assume $f, g \notin \p'$.
    Write $f = \sum_d f_d$ and $g = \sum_d g_d$ with $f_d, g_d \in S_d$.
    So
    \begin{equation*}
        \sum_{i, j} f_i g_j = \sum_n \sum_{i + j = n} f_i g_j \in \p'
    \end{equation*}
    Since $\p'$ is homogeneous, it follows that
    \begin{equation*}
        \sum_{i + j = n} f_i g_j \in \p'
    \end{equation*}
    for all $n \in \Z$.
    Let now $d$ resp. $e$ be maximal such that $f_d \notin \p'$ resp. $g_e \notin \p'$.
    We have
    \begin{equation*}
        f_d g_e + \sum_{\substack{i + j = d + e\\(i, j) \neq (d, e)}} \underbrace{f_i g_j}_{\in \p'} = \sum_{i + j = d + e} f_i g_j \in \p'
    \end{equation*}
    and so $f_d g_e \in \p' \subseteq \p$.
    Since $\p$ is prime, it follows that $f_d \in \p$ or $g_e \in \p$.
    However both $f_d$ and $g_e$ are homogeneous, so $f_d \in \p'$ or $g_e \in \p'$, a contradiction.
\end{proof}
\begin{lemma}[``Covering Trick'']
    \label{prop:covering_trick}
    Let $f, f_i \in S_+$ be homogeneous.
    Then $\D{f} \subseteq \bigcup_i \D{f_i}$ if and only if $f^n \in \langle f_i \ | \ i \rangle$ for some $n \in \N$
\end{lemma}
\begin{proof}
    The direction $\Leftarrow$ is clear.
    So assume $\D{f} \subseteq \bigcup_i \D{f_i}$.
    If no $f^n$ is in $\a := \langle f_i \ | \ i \rangle$, then $f$ is not nilpotent in $S/\a$ and so $(S/\a)_f$ is not the zero ring.
    Hence, $(S/\a)_f$ has a prime ideal, and its preimage under $S \to S/\a \to (S/\a)_f$ yields a prime ideal $\p'$ containing $\a$ and not containing $f$.
    Since $f \in S_+$ note that $S_+ \not\subseteq \p'$ and as both $f$ and $\a$ are homogeneous, Lemma~\ref{prop:homogeneous_part_prime} yields now that $\langle g \in \p' \ | \ \text{$g$ homogeneous} \rangle$ is a prime ideal in $\D{f}$ but not in $\bigcup_i \D{f_i}$, contradicting the assumption.
\end{proof}
\begin{corollary}
    \label{prop:basic_set_inclusion_divisibility}
    If $\D{g} \subseteq \D{f}$ with $f, g \in S_+$, then there is a homogeneous $h \in S$ such that $g^n = fh$ for some $n \in \N$.
\end{corollary}
\begin{proof}
    By assumption, $\D{f}$ covers $\D{g}$, thus by the Covering Trick (Lemma~\ref{prop:covering_trick}), observe that $g^n \in \langle f \rangle$ for some $n \in \N$.
    The claim follows.
\end{proof}
\begin{prop}
    Let $B = \{ \D{f} \ | \ \text{$f \in S_+$ homogeneous}\}$.
    The functor\footnote{The map on arrows is well-defined by Corollary~\ref{prop:basic_set_inclusion_divisibility}.}
    \begin{align*}
        \mathcal{F}: \restr{\Top(\Proj(S))}{B} \to \Ring, \quad \D{f} &\mapsto (S_f)_0 \\
        (\D{fg} \subseteq \D{f}) &\mapsto \Bigl( \restr{\cdot}{\D{fg}}: \frac s {f^n} \mapsto \frac {sg^n} {(fg)^n} \Bigr)
    \end{align*}
    is a B-sheaf on $B$ (here $\Top(X)$ is the category given by the open sets of $X$ and their inclusion, as defined in the lecture).
\end{prop}
\begin{proof}\footnote{I guess I have included the proof mostly for the sake of completeness - apart from presentational issues, it is just copied from the lecture notes.}
    Clearly, $\mathcal{F}$ is a functor and thus a presheaf.
    Hence, we have to show the local-to-global property.

    Let $\D{f} = \bigcup_{i \in I} \D{g_if}$ be a cover and $s_i \in \mathcal{F}(\D{g_if})$ such that
    \begin{equation*}
        \forall x \in \D{g_if} \cap \D{g_jf} \ \exists V \in B: \ V \subseteq \D{g_if} \cap \D{g_jf}, \ x \in V, \ \frac {s_i} 1 = \frac {s_j} 1 \in \mathcal{F}(V)
    \end{equation*}
    We show that the $s_i$ glue to a unique $s \in \mathcal{F}(\D{f})$.
    To show uniqueness, assume there are $\alpha/f^N, \beta/f^N \in \D{f}$ with
    \begin{equation*}
        \restr{\frac {\alpha} {f^N}}{\D{g_if}} = \restr{\frac {\beta} {f^N}}{\D{g_if}} \quad \text{for all $i$}
    \end{equation*}
    Therefore there is $n_i \in \N$ such that
    \begin{equation*}
        (f g_i)^{n_i} (\alpha f^N + \beta f^N) = 0
    \end{equation*}

    wlog there are only finitely many $i$ (by the Covering Trick, Lemma~\ref{prop:covering_trick}).

    wlog $N$ is sufficiently large (we can always increase $N$), and so $(fg_i)^N(\alpha - \beta) = 0$.

    Since $\bigcup_i \D{(fg_i)^N} = \bigcup_i \D{fg_i} = \D{f}$ it follows that $f^n \in \langle (fg_i)^N \ | \ i \rangle$ for some $n \in \N$ (again the Covering Trick, Lemma~\ref{prop:covering_trick}).
    Therefore we find
    \begin{equation*}
        f^n(\alpha - \beta) \in \langle (f g_i)^N \ | \ i \rangle (\alpha - \beta) = \langle (f g_i)^N (\alpha - \beta) \ | \ i \rangle = \{ 0 \}
    \end{equation*}
    and so $\alpha = \beta \in S_f$.

    Now we show existence.
    By the uniqueness above, it follows that
    \begin{equation*}
        \restr{s_i}{\D{fg_ig_j}} = \restr{s_i}{\D{fg_i} \cap \D{fg_j}} = \restr{s_j}{\D{fg_i} \cap \D{fg_j}} = \restr{s_j}{\D{fg_ig_j}}
    \end{equation*}
    wlog have again a finite cover, i.e. only finitely many $g_i$.
    Hence find an $N \in \N$ such that each $s_i = s'_i / (fg_i)^N$ with $s'_i \in S$ homogeneous.
    By possibly replacing $N$ with a bigger $N$, we can now assume that
    \begin{equation*}
        (f^2g_ig_j)^N \left( s'_i (f g_j)^N - s'_j (f g_i)^N \right) = 0 \quad \text{as $\restr{s_i}{\D{fg_ig_j}} = \restr{s_j}{\D{fg_ig_j}}$}
    \end{equation*}
    Now note that
    \begin{equation*}
        s_i = \frac {a_i} {b_i} \quad \text{with}\ a_i = s'_i(f g_i)^N, \ b_i = (f g_i)^{2N}
    \end{equation*}
    and
    \begin{equation*}
        a_i b_j - a_j b_i = s'_i (f g_i)^N (f g_j)^{2N} - s'_j (f g_j)^N (f g_i)^{2N} = \underbrace{(f^2g_ig_j)^N \left(s'_i (fg_j)^N - s'_j (fg_i)^N \right)}_{= 0}
    \end{equation*}
    Now observe that $\D{b_i} = \D{fg_i}$ and so $f^n \in \langle b_i \ | \ i \rangle$ for some $n$ (once again the Covering Trick, Lemma~\ref{prop:covering_trick}).
    Let $f^n = \sum_i r_i b_i$ and get
    \begin{equation*}
        a_i f^n = \sum_l r_l b_l a_i = \sum_l r_l a_l b_i = b_i \sum_l r_l a_l
    \end{equation*}
    Note that $a_i, b_i, f$ are homogeneous, and so we can also choose $r_i$ to be homogeneous.
    Then find that $0 = \deg(s_i) = \deg(a_i) - \deg(b_i) = \deg(\sum r_l a_l) - n\deg(f)$.

    Thus
    \begin{equation*}
        s_i = s := \frac {\sum_l r_l a_l} {f^n} \in S_f
    \end{equation*}
    and since $\deg(\sum r_l a_l) = n\deg(f)$ we find that $s \in (S_f)_0 = \mathcal{F}(\D{f})$.
    Clearly $\restr{s}{\D{fg_i}} = s_i$ and the claim follows.
\end{proof}
Now the B-sheaf extension theorem from the lecture (from Section 1.11) yields the following corollary.
\begin{corollary}
    The B-sheaf $\mathcal{F}$ can be (uniquely, up to isomorphism) extended to a sheaf $\O_{\Proj(S)}$ on $\Proj(S)$.
\end{corollary}
The next to lemmas are based on \cite[Prop. II.2.5]{hartshorne} and will be used to deduce that $(\Proj(S), \O_{\Proj(S)})$ is a scheme.
\begin{lemma}
    For $\p \in \Proj(S)$, the stalk
    \begin{equation*}
        \O_{\Proj(S), \p} = (T^{-1}S)_0
    \end{equation*}
    is a local ring, where $T = \{ f \notin \p \ | \ \text{$f$ homogeneous}\}$ contains all homogeneous elements not in $\p$.
\end{lemma}
\begin{proof}
    Consider the ideal
    \begin{equation*}
        \m = \Bigl\{ \frac f g \in (T^{-1}S)_0 \ \Bigm| \ f \in \p \Bigr\}
    \end{equation*}
    We claim this is the unique maximal ideal of $(T^{-1}S)_0$.

@@

            \node (Sf) at (8.5, 0) {$(S[x_0, ..., x_n]_{x_i})_0$};
            \node (Sg) at (5, -1.5) {$(S[x_0, ..., x_n]_{x_j})_0$};
            \node (Sfg) at (8.5, -3) {$(S[x_0, ..., x_n]_{x_ix_j})_0$};

@@

            \draw [->] (Rf) -- (Rfg);
            \draw [->] (Rg) -- (Rfg);
            \draw [->] (Sf) -- (Sfg);
            \draw [->] (Sg) -- (Sfg);
        \end{tikzpicture}
    \end{center}
    commutes, and so the restrictions of $\phi_i$ resp. $\phi_j$ to $\D{x_ix_j}$ are both induced by the homomorphism $\iota_{x_ix_j}$, hence they are equal.
    The claim follows.
\end{proof}
This lemma also justifies the usual definition of projective $n$-space over a scheme $B$ as
\begin{equation*}
    \P_B^n := B \times_{\Spec(\Z)} \P_\Z^n
\end{equation*}

@@

    It is not hard to show that the morphism is unique\footnote{In fact, the statement that there is at most one such morphism is equivalent to $\P_R^n \to \Spec(R)$ being separated (by \citestacks{01L0}), so follows from Proposition~\ref{prop:projective_space_separated}.}, and the claim follows.
\end{proof}
Note that in some way, this is the same proof as before.
The core idea - namely the choice of the smallest element $g_{k0}$ (w.r.t. divisibility by $R$) - more or less achieves the same effect as the more complicated ideal argument from Lemma~\ref{prop:adding_coeff_ideal_does_not_introduce_irrelevant_ideal}.
\printbibliography
\end{document}

>>>>

In this miniproject, we want to study the $\Proj$-construction in the theory of schemes, that can be used to associate a scheme given a graded ring.
In particular, it is very closely related to projective space (hence the name).

We begin the work by defining the scheme $\Proj(S)$ for a graded ring $S$, and showing that it is well-defined.
Furthermore, we study some very basic properties.
Then we focus on the case that our graded ring $S$ is an $R$-algebra, in particular, we are interested in the case of the polynomial ring $R[x_0, ..., x_n]$.
This case is important, since it yields the projective space $\P_R^n$.

Finally, we use the theory developed in the work to prove the main result, namely that projective space is a proper scheme.
We present two proofs of this, one relatively short and concrete (but complicated) one, relying on commutative algebra.
The other proof is the standard proof, as presented in works like \cite{hartshorne}.
For the sake of brevity, we will only present parts of this proof, but try to give a little bit of intuition.

@@

    Consider $\p \in \D{f} \cap \D{g}$, so $f, g \notin \p$.
    Since $\p$ is prime, it follows that $fg \notin \p$ and so $\D{fg} \subseteq \D{f} \cap \D{g}$ is an open neighborhood of $\p$.
\end{proof}
\begin{lemma}
    \label{prop:homogeneous_part_prime}
    Let $\p \leq S$ be a prime ideal.
    Then
    \begin{equation*}
        \p' := \langle f \in \p \ | \ \text{$f$ homogeneous} \rangle \leq S
    \end{equation*}
    is a (homogeneous) prime ideal.
\end{lemma}
\begin{proof}
    Consider $f, g \in S$ with $f g \in \p'$ and assume $f, g \notin \p'$.
    Write $f = \sum_d f_d$ and $g = \sum_d g_d$ with $f_d, g_d \in S_d$.
    So
    \begin{equation*}
        \sum_{i, j} f_i g_j = \sum_n \sum_{i + j = n} f_i g_j \in \p'
    \end{equation*}
    Since $\p'$ is homogeneous, it follows that
    \begin{equation*}
        \sum_{i + j = n} f_i g_j \in \p'
    \end{equation*}
    for all $n \in \Z$.
    Let now $d$ resp. $e$ be maximal such that $f_d \notin \p'$ resp. $g_e \notin \p'$.
    We have
    \begin{equation*}
        f_d g_e + \sum_{\substack{i + j = d + e\\(i, j) \neq (d, e)}} \underbrace{f_i g_j}_{\in \p'} = \sum_{i + j = d + e} f_i g_j \in \p'
    \end{equation*}
    and so $f_d g_e \in \p' \subseteq \p$.
    Since $\p$ is prime, it follows that $f_d \in \p$ or $g_e \in \p$.
    However both $f_d$ and $g_e$ are homogeneous, so $f_d \in \p'$ or $g_e \in \p'$, a contradiction.
\end{proof}
\begin{lemma}[``Covering Trick'']
    \label{prop:covering_trick}
    Let $f, f_i \in S_+$ be homogeneous.
    Then $\D{f} \subseteq \bigcup_i \D{f_i}$ if and only if $f^n \in \langle f_i \ | \ i \rangle$ for some $n \in \N$
\end{lemma}
\begin{proof}
    The direction $\Leftarrow$ is clear.
    So assume $\D{f} \subseteq \bigcup_i \D{f_i}$.
    If no $f^n$ is in $\a := \langle f_i \ | \ i \rangle$, then $f$ is not nilpotent in $S/\a$ and so $(S/\a)_f$ is not the zero ring.
    Hence, $(S/\a)_f$ has a prime ideal, and its preimage under $S \to S/\a \to (S/\a)_f$ yields a prime ideal $\p'$ containing $\a$ and not containing $f$.
    Since $f \in S_+$ note that $S_+ \not\subseteq \p'$ and as both $f$ and $\a$ are homogeneous, Lemma~\ref{prop:homogeneous_part_prime} yields now that $\langle g \in \p' \ | \ \text{$g$ homogeneous} \rangle$ is a prime ideal in $\D{f}$ but not in $\bigcup_i \D{f_i}$, contradicting the assumption.
\end{proof}
\begin{corollary}
    \label{prop:basic_set_inclusion_divisibility}
    If $\D{g} \subseteq \D{f}$ with $f, g \in S_+$, then there is a homogeneous $h \in S$ such that $g^n = fh$ for some $n \in \N$.
\end{corollary}
\begin{prop}
    Let $B = \{ \D{f} \ | \ \text{$f \in S_+$ homogeneous}\}$.
    The functor\footnote{The map on arrows is well-defined by Corollary~\ref{prop:basic_set_inclusion_divisibility}.}
    \begin{align*}
        \mathcal{F}: \restr{\Top(\Proj(S))}{B} \to \Ring, \quad \D{f} &\mapsto (S_f)_0 \\
        (\D{fg} \subseteq \D{f}) &\mapsto \Bigl( \restr{\cdot}{\D{fg}}: \frac s {f^n} \mapsto \frac {sg^n} {(fg)^n} \Bigr)
    \end{align*}
    is a B-sheaf on $B$ (here $\Top(X)$ is the category given by the open sets of $X$ and their inclusion, as defined in the lecture).
\end{prop}
\begin{proof}
    This is exactly the proof from the lecture, Section 1.12.
    Note that we have all necessary ingredients, in particular the Covering Trick~\ref{prop:covering_trick}.
    Hence, the only thing to note is that the gluing produces a homogeneous global section if one starts with homogeneous local sections.
    It is easy to check that this is indeed the case.
\end{proof}
Now the B-sheaf extension theorem from the lecture (from Section 1.11) yields the following corollary.
\begin{corollary}
    The B-sheaf $\mathcal{F}$ can be (uniquely, up to isomorphism) extended to a sheaf $\O_{\Proj(S)}$ on $\Proj(S)$.
\end{corollary}
The next to lemmas are based on \cite[Prop. II.2.5]{hartshorne} and will be used to deduce that $(\Proj(S), \O_{\Proj(S)})$ is a scheme.
\begin{lemma}
    For $\p \in \Proj(S)$, the stalk
    \begin{equation*}
        \O_{\Proj(S), \p} = (T^{-1}S)_0
    \end{equation*}
    is a local ring, where $T = \{ f \notin \p \ | \ \text{$f$ homogeneous}\}$ contains all homogeneous elements not in $\p$.
\end{lemma}
\begin{proof}
    Consider the ideal
    \begin{equation*}
        \m = \Bigl\{ \frac f g \in (T^{-1}S)_0 \ \Bigm| \ f \in \p \Bigr\}
    \end{equation*}
    We claim this is the unique maximal ideal of $(T^{-1}S)_0$.

@@

            \node (Sf) at (7.5, 0) {$(S[x_0, ..., x_n]_{x_i})_0$};
            \node (Sg) at (5, -1.5) {$(S[x_0, ..., x_n]_{x_j})_0$};
            \node (Sfg) at (7.5, -3) {$(S[x_0, ..., x_n]_{x_ix_j})_0$};

@@

            \draw [->] (Rf) -- (Rfg);
            \draw [->] (Rg) -- (Rfg);
            \draw [->] (Sf) -- (Sfg);
            \draw [->] (Sg) -- (Sfg);
        \end{tikzpicture}
    \end{center}
    commutes, and so the restrictions of $\phi_i$ resp. $\phi_j$ to $\D{x_ix_j}$ are both induced by the homomorphism $\iota_{x_ix_j}$, hence they are equal.
    The claim follows.
\end{proof}
This lemma also justifies the usual definition of \emph{projective $n$-space over a scheme $B$} as
\begin{equation*}
    \P_B^n := B \times_{\Spec(\Z)} \P_\Z^n
\end{equation*}

@@

    It is not hard to show that the morphism is unique\footnote{In fact, the statement that there is at most one such morphism is equivalent to $\P_R^n \to \Spec(R)$ being separated (by \citestacks{01L0}), so follows from Proposition~\ref{prop:projective_space_separated}.}, and the claim follows.
\end{proof}
Note that in some way, this is the same proof as before.
The core idea - namely the choice of the smallest element $g_{k0}$ (w.r.t. divisibility by $R$) - more or less achieves the same effect as the more complicated ideal argument from Lemma~\ref{prop:adding_coeff_ideal_does_not_introduce_irrelevant_ideal}.

\section*{Conclusion}
In this work, we have seen how to construct the scheme $\Proj(S)$ given a graded ring $S$.
Furthermore, we have peeked into its properties, and used it to prove an important statement about projective space.
Additionally, we have tried to explain a little bit what is going on in the proof, and compared two different versions of proof.
The first proof is my own work\footnote{Meaning I could find it nowhere online. However, I do not believe that it is original, and in some sense, it relies on the same idea as the standard proof.}, while the second proof is the standard proof found in many textbooks.

If one continues that line of thought, the next step would probably be to study \emph{projective morphisms}, i.e. morphisms that factor through projective space via a closed immersion, and show that they are always proper.
After that, it seems like an interesting idea to examine the phenomenon of properness closer.
For example \cite{hartshorne} mentioned that not all proper morphisms are projective, so it might be instructive to find examples for that and study in which other situations one gets a projective morphism.

Finally, another line of continuing the work would be to study the cohomology of $\Proj(S)$, and how it relates to the ring $S$.
From the lecture, we already know that projective space $\P_\Z^n$ has a first \v{C}ech-cohomology group isomorphic to $\Z$.
Starting from that, it would be interesting to see what the cohomology group of $\Proj(S)$ for other graded rings $S$ is.
A natural first step would be to consider $S = R[x_0, ..., x_n]/\a$ for a homogeneous ideal $\a$, i.e. study the cohomology of closed subschemes of $\P_R^n$.
This can already become quite complicated, as for example the case of Elliptic Curves shows.
In the end, it will be interesting if the representation of a scheme as $\Proj(S)$ gives us better access to computing cohomology as the standard way to cover it by affine opens.

\printbibliography
\end{document}
